Unveiling Vulnerabilities

### 2024-04-24 paper/paper.qmd
saving readme and paper pdf + qmd

---
title: "Unveiling Vulnerabilities"
subtitle: "An Analysis of Causes of Death Across Age Groups and Genders Within Toronto's Homelessness Population"
author: "Faiza Imam"
thanks: "Code and data are available at: https://github.com/fazim07/deaths_gender"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

This paper aims to identify the correlation between gender and age within the homeless population in Toronto from the year 2017 - 2023, specifically examining how age and gender intersect in terms of homelessness experiences and demographic characteristics. In examining the data set concerning causes of death, including homelessness as a potential factor, I hypothesize that there is a notable relationship between female homelessness and mortality rates across different age groups. Specifically, the anticipation that younger age groups (<20 and 20-39) among homeless females may exhibit higher mortality rates compared to older age groups (40-59 and 60+), potentially due to factors such as exposure to harsh living conditions, limited access to healthcare, and increased vulnerability to external threats.

@@

The original data set contained six variables, for this paper, I have selected five of the six. The first variable, *id* contains an id number which is associated with the following variables, *cause of death*, *age group*, *year of death*, and *gender*.  There are a total of 253 entries. The *cause of death* variable is listened from 12 categories. These categories include homicide, accident, cardiovascular disease, cancer, COVID-19, drug toxicity, pneumonia, suicide, others and unknown. There are no notices nor indications on what the category other and unknown classifies as. The *age group* variable is a set of four age ranges, <20 years old, 20-39 years of age, 40-59 years of age and 60+ years of age @tbl-Agegroup. *Year of death*, covers the years from 2017 to 2023. Lastly, the gender variable houses, three categories, male, female, and unknown @tbl-Gender. Once again, the *unknown* has not been classified.

@@

# Group by Cause_of_death and Age_group, maintaining the age order
AgevCause <- home_data %>%
  group_by(Cause_of_death, Age_group) %>%
  summarize(total_deaths = sum(death_toll))

@@

```{r}
#| label: genderyear
#| fig-cap: "Total deaths by age group from 2017-2023"
#| fig-subcap: ["Total Deaths by Age in 2017", "Total Deaths by Age in 2018", "Total Deaths by Age in 2019", "Total Deaths by Age in 2020", "Total Deaths by Age in 2021","Total Deaths by Age in 2022","Total Deaths by Age in 2023"]
#| layout-ncol: 2
#| echo: false
#| warning: false
#| message: false

@@

```
# Results

@@

## Causes of Death by Age Group

## Causes of Death by Gender

>>>>

---
title: "Unveiling Vulnerabilities"
subtitle: "An Analysis of Causes of Death Across Age Groups and Genders Within Toronto's Homelessness Population"
author: "Faiza Imam"
thanks: "Code and data are available at: https://github.com/fazim07/deaths_gender"
date: today
date-format: long
abstract: "This paper aims to explore the relationship between male homelessness and mortality rates across age groups, focusing on causes of death related to homelessness in Toronto from the years 2017- 2023. The hypothesis suggests that younger homeless males (<20 and 20-39) may have higher mortality rates due to harsh living conditions, limited healthcare access, and increased vulnerability. The results indicate that while male homeless mortality rates are generally higher than reported female death tolls, the most affected age demographic is the 40-59 age range rather than the hypothesized age group."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

This paper aims to identify the correlation between gender and age within the homeless population in Toronto from the year 2017 - 2023, specifically examining how age and gender intersect in terms of homelessness experiences and demographic characteristics. In examining the data set concerning causes of death, including homelessness as a potential factor, I hypothesize that there is a notable relationship between male homelessness and mortality rates across different age groups. Specifically, the anticipation that younger age groups (<20 and 20-39) among homeless males may exhibit higher mortality rates compared to older age groups (40-59 and 60+), potentially due to factors such as exposure to harsh living conditions, limited access to healthcare, and increased vulnerability to external threats.

@@

The original data set contained six variables, for this paper, I have selected five of the six. The first variable, *id* contains an id number which is associated with the following variables, *cause of death*, *age group*, *year of death*, and *gender*.  There are a total of 253 entries. The *cause of death* variable is listened from 12 categories. These categories include homicide, accident, cardiovascular disease, cancer, COVID-19, drug toxicity, pneumonia, suicide, others and unknown. There are no notices nor indications on what the category other and unknown classifies as. *Year of death*, covers the years from 2017 to 2023. Lastly, the gender variable houses, three categories, male, female, and unknown @tbl-Gender. Once again, the *unknown* has not been classified. We can observe by the table that Male homeless victims have highest death tolls sitting at 749, while female victims come in second with 210, followed by unknown indivudals who are accounted for as 3 The *age group* variable is a set of four age ranges, <20 years old, 20-39 years of age, 40-59 years of age and 60+ years of age @tbl-Agegroup. We can see that there are 9 deaths reported for the age gorup under 20, followed by the 20-39 age range, then the 60+ age range and finally, the 40-59 age range having the highest death tolls among each demograph.

@@

# Group by Cause_of_death and Age_group, maintaining the age order
AgevCause <- home_data %>%
  group_by(Cause_of_death, Age_group) %>%
  summarize(total_deaths = sum(death_toll))
AgevCause

@@

```{r}
#| label: genderyear
#| fig-cap: "Total Deaths by Gender from 2017-2023"
#| fig-subcap: ["Total Deaths by Gender in 2017", "Total Deaths by Gender in 2018", "Total Deaths by Gender in 2019", "Total Deaths by Gender in 2020", "Total Deaths by Gender in 2021","Total Deaths by Gender in 2022","Total Deaths by Gender in 2023"]
#| layout-ncol: 2
#| echo: false
#| warning: false
#| message: false

@@

# Results

## Age Groups and Cause of Death

## Gender and Cause of Death

@@

## Male to Female Cause of Death Trends

## Age Demograph Cause of Death Trends

## Male Age Specific Death Trends
